---
title: "Introduction to urbanheatisland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to urbanheatisland}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`urbanheatisland` supports urban heat island / biometeorology workflows that
combine:

1. **Land Surface Temperature (LST)**, from MODIS (via the NASA AppEEARS
   API) or from local files you already have.
2. **Interpolated air temperature**, from station observations, using the
   same family of methods (IDW, kriging, thin-plate spline, TIN, nearest
   neighbour) used in the companion `R.IBE.airqino` package.
3. **Urban heat island metrics and downscaling**, combining the two.
4. **Local spatial hotspot analysis** (Getis-Ord Gi\*, Local Moran's I) on
   any of the resulting rasters.

## 1. Getting LST

### From NASA AppEEARS (MODIS MOD11A2, 8-day 1km composite)

```{r}
library(urbanheatisland)

Sys.setenv(EARTHDATA_USER = "your_username")
Sys.setenv(EARTHDATA_PASS = "your_password")

lst <- get_modis_lst(
  aoi        = my_aoi_polygon,   # sf polygon, EPSG:4326
  start_date = "06-01-2026",
  end_date   = "08-31-2026"
)
```

This logs in, submits an area task, polls until it's done, downloads the
GeoTIFFs, applies the MODIS scale factor and Kelvin-to-Celsius conversion,
and masks low-quality pixels using the QC layer — all in one call. Each
step is also available individually (`lst_appeears_login()`,
`lst_appeears_submit_task()`, `lst_appeears_wait()`,
`lst_appeears_download()`) if you need more control, e.g. to submit several
tasks before waiting on any of them.

### From local files

```{r}
lst <- read_lst_local(
  "MOD11A2.A2026153.h18v04.061.LST_Day_1km.tif",
  qc_path = "MOD11A2.A2026153.h18v04.061.QC_Day.tif"
)
```

`decode_modis_qc()` exposes the QC-bit decoding directly if you want to
build a custom mask, and `mosaic_lst()` / `composite_lst()` help combine
multiple tiles or dates (e.g. cloud-gap filling across an 8-day window).

## 2. Interpolating air temperature

```{r}
air_temp <- interpolate_airtemp(
  stations   = station_points,   # sf POINT, with a temperature column
  value_col  = "temp_c",
  template   = lst,              # defines the output grid
  method     = "kriging"
)
```

If you're not sure which method suits your station network, compare a few
with leave-one-out cross-validation:

```{r}
cv_interpolate_airtemp(station_points, "temp_c", template = lst,
                        methods = c("idw", "kriging", "nn"))
```

## 3. Urban heat island intensity and downscaling

```{r}
uhi <- compute_uhi(lst, rural_ref = rural_reference_polygon)

# Downscale coarser air temperature to LST's finer resolution
air_temp_fine <- downscale_airtemp(air_temp, lst, method = "regression")

# Or with regression-kriging of the residuals (uses station data directly)
air_temp_fine_rk <- downscale_airtemp(
  air_temp, lst,
  stations = station_points, value_col = "temp_c",
  method = "rk"
)
```

## 4. Hotspot analysis

```{r}
gi <- gi_star_raster(lst, d = 1500)   # ~1.5 km neighbourhood
hotspots <- classify_hotspots(gi, apply_fdr = TRUE)
plot_hotspots(hotspots)

lisa <- local_moran_raster(lst, d = 1500)  # HH / LL / HL / LH clusters
```

`gi_star_raster()` automatically switches from an exact `spdep`-based
computation to a faster `terra::focal`-based approximation above
`max_cells_direct` valid cells (default 200,000) — tune this for large
rasters.

## A note on false discovery rate

With thousands of raster cells being tested simultaneously,
`classify_hotspots(apply_fdr = TRUE)` (the default) applies a
Benjamini-Hochberg FDR correction to the Gi\* p-values before classifying
hot/cold spots, reducing the chance of spurious clusters compared to raw
z-score cutoffs.
